---
title: "Stats 306: Lecture 3"
subtitle: "More data exploration with ggplot"
author: "Mark Fredrickson"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
```

## Review: More on R

* Learned more about R's functions (required and optional arguments)
* Piping and assignment
* Special values: `NA`, `NULL`, and `NaN`
* Vectors: ordered collections of same type, many computations are *vectorized*
* Lists: ordered collections of mixed types
* Tables: lists of vectors of the same length (column oriented)
* Other items: indexing with `[]` and `$`, `dim` and `colnames` for tables

## Vectorized computations in R

In R, since (almost) *everything* is a vector (of length 1 if need be), most computations in R allow the use of **vectorized** computations: computations that apply to **each item in a vector** or **use the entire vector**.

```{r}
miles_per_gallon_city <- mpg$cty
sum(miles_per_gallon_city)
gallons_per_mile_city <- 1 / miles_per_gallon_city
miles_per_gallon_city[1:3]
gallons_per_mile_city[1:3]
total_mpg <- miles_per_gallon_city + mpg$hwy
mpg$hwy[1:3]
total_mpg[1:3]
```

## Vectorized Computations with Logical Values

We saw relational operators in the last lecture (`<`, `>`, `==`, etc.) and that they can apply in a vectorized way.

```{r}
top_half <- mpg$cty > median(mpg$cty)
top_half[1:5]

## average highway efficiency of cars in top 50% of city efficiency:
mpg$hwy[top_half] |> mean()
```

## Other uses for logicals

R has a convenient convention that `TRUE == 1` and `FALSE == 0`. Notice that if we want to **count** this is equivalent to **summing** a logical vector.

```{r}
# How many cars get more than 30 miles per gallon in the city?
sum(mpg$cty > 30)
```

## Exercise

How many diamonds in the `diamonds` table cost more than $5000? What was the mean carat rating of these diamonds? (Bonus: what *proportion* of diamonds cost more than $5000 -- can you figure it out with one function call?)

```{r expensive-diamonds, exercise = TRUE}

```


```{r expensive-diamonds-solution}
expensive <- diamonds$price > 5000
sum(expensive) ## how many
mean(keep(diamonds$price, expensive)) ## could also use [] indexing
mean(expensive) ## bonus
```



## Review: Graphing and Plotting

* Semiology of Graphics: Describes fundamental aspects of graphs, tasks
  * Visual variables (Position, Size, Value, Texture, Color, Orientation, Shape) 
  * Relationships/Tasks: Selection (grouping), association (similarity across groups), order, quantity
* Grammar of Graphics, `ggplot2` library

## The components of a ggplot graph

>* The graph object itself (creating using `ggplot(data)`)
>* A set of *aesthetic* mappings (connecting data to visual variables)
>* Layers: collections of geometric elements (`geom_*()`) and statistical transformations (`stat_*()`)
>* Scales: information on the range or composition of variables
>* Coordinate systems: how the data are arranged spatially
>* Facet: breaking a single plot into many, similar plots
>* Theme: all the other color and printing aspects of the plot

## Statistical Summaries

* In addition to the raw data (or our calculations), many plots use **data summaries** 
* `ggplot2` calls these **summary statistics** or `stat_*` functions
* We already saw a summaries in the boxplot: quantiles, twice IQR bars
* We can access summaries that geometries compute and add additional summaries.

## Boxplot as statistic

```{r}
median(mpg$displ)
quantile(mpg$displ, c(0.25, 0.75))
```

```{r}
ggplot(data = mpg,
       aes(x = displ)) + 
  geom_boxplot() 
```


## Adding computed summaries

The `stat_summary` function allows you to use any function to summarize 

```{r}
ggplot(data = mpg,
       aes(x = displ, y = class)) + 
  geom_boxplot() +
  stat_summary(fun = mean, size = 3, color = "red", geom = "point") 
```

## Trend lines

When using scatter plots, one of the most common summaries is a **trend line**.
```{r}
ggplot(data = mpg,
       aes(x = displ, y = hwy)) + 
  geom_point(position = "jitter", alpha = 0.25) +
  stat_smooth() # geom_smooth also works
```

## More layering

```{r}
ggplot(data = mpg, aes(x = displ)) +
  geom_point(aes(y = hwy), color = "orange") +
  geom_point(aes(y = cty), color = "blue") +
  stat_smooth(aes(y = hwy), lty = 1, color = "black") +
  stat_smooth(aes(y = cty), lty = 2, color = "black")
```

We'll see a better way to make this table when we talk about tall vs. wide format data.

## Overriding defaults of `stat` functions

Each `geom_*` has a default statistic function. We can override this.

```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar() # default stat is count
```

## Average highway efficiency by class

```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  geom_bar(stat = "summary", fun = "mean") 
```

## Replacing tables

We often use **tables** in documents to give numerical summaries. But why not
replace those with a nice graphic?
```{r}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  stat_summary(
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

## Exporting Plots

* Automatically included in knitted RMarkdown documents
* Export from the Plots panel in RStudio/Posit Cloud
* `ggsave` function to save to file (variety of formats)

