---
subtitle: "Stats 306: Lecture 3"
title: "Statistical Programming"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
```


## R: a language for samples

Recall that R is a **statistical** programming language. In statistics we often work with:

>* **Samples**, composed of individual observations
>* **Variables**, something measured on each unit in the sample

Tabular data: columns are variables, rows are observations

## Lists and Vectors

* A **vector** is an ordered collection of the same type of data. Created using the `c` function (concatenate).
* A **list** is an ordered collection of (potentially) different types of data. Created using the `list` function.
* Both can have **names** to indicate individual items.
* Lists can serve a similar role to dictionaries and hash tables in other languages.

## Vectors, basic

```{r}
phone_digits <- c(8, 6, 7, 4, 3, 0, 9)
phone_plus_area_code <- c(7, 3, 4, phone_digits)
length(phone_plus_area_code)
sum(phone_plus_area_code)
```

## Vectors, indexing

The `runif` function will create randomly generated values in the range (0, 1).

```{r}
random_u01 <- runif(10) # create 10 values
print(random_u01)
random_u01[1] # first item
random_u01[10] # last item
random_u01[-1] # all but first
random_u01[1:3] # first three
random_u01[3:6] # a "slice" in the middle
random_u01[c(1, 3, 7, 1)] # arbitrary choices
```


## Exercise

Pull out the last three items in the vectors `letters`. A little help to get you started:

```{r lettersxyz, exercise = TRUE}
length(letters)
```


```{r lettersxyz-solution}
letters[c(24, 25, 26)]
letters[-(1:23)]
```

## Relational Operators

```{r}
x <- 1
y <- 2

x < y
x > y
x <= y
x >= y
x == y
x != y

!(x == y) # negation
```

## Vectorized computations

In the previous slide we compared two *scalar values* (i.e., single numbers). The result was a scalar *logical* value.

```{r}
x <- 3
y <- 10
x > y
```

We can also compare two vectors of the same length.

```{r}
x <- c(1, 2, 3)
y <- c(10, -1, 3)

x < y
```

We call operations on collections of data like this *vectorized computations.*

## Other vectorized computations

We can also operate on vectors with vectorized computations using airthmetic operators.

```{r}
x <- c(1, 2, 3)
y <- c(10, -1, 3)
x + y
2 * x
```

Notice that we multiplied a vector of length 1 (`2`) by a vector of length 3 (`x`). This is called *recycling* and is a common feature of R.

```{r}
# 2 * x is the same as:
c(2, 2, 2) * x
```

Be careful if you see this warning message:

```{r}
c(2, 2) * c(3, 3, 3)
```

It means to tried to operate on two lists of different sizes where one could not be evenly recycled to match the other. Usually, this indicates a programming error.


## Lists, heterogenous collections

Vectors must be the same type, lists can be **heterogeneous**:
```{r}
list("string", 7, 1:4)
```

## Named collections

Vectors can have names (`c(first = 1, second = 22, third = 333)`), but names for lists are more common.
```{r}
named_list <- list(a_string = "A string", a_vector = 1:4)
named_list[["a_string"]] # note double brackets
named_list$a_vector
```

## Tables: Lists of Vectors

* In a **sample**, we suppose we have $k$ variables measured for each of $n$ units.
* For each variable, place the $n$ values in a vector.
* Collect all such vectors into a list of length $k$.
* Result: $n \times k$ table

## Tables in R

There are a few different types of tables in R (`data.frame`, `tibble`). We will treat them interchangeably for the most part.

Here are some of the most common operations we will run on tables:

```{r}
class(mpg)
dim(mpg)
colnames(mpg)
head(mpg)
summary(mpg)
```

## Accessing individual columns

Recall, the `$` operator lets us get items from lists:

```{r}
mpg$hwy |> mean() # equiv to mean(mpg$hwy)
```

## Functions for basic descriptive statistics

```{r}
sum(mpg$hwy)
mean(mpg$hwy)
var(mpg$hwy)
sd(mpg$hwy) 
median(mpg$hwy)
IQR(mpg$hwy)
quantile(mpg$hwy, c(0.25, 0.75))
```

## Exercise

What is the average `year` of the cars in `mpg`? What is the standard deviation of the city mileage (`cty`)?

```{r mpgsummary, exercise = TRUE}

```

```{r mpgsummary-solution}
mean(mpg$year)
mpg$cty |> sd()
```

## Vectorized computations with columns

Since the columns of a table are *vectors*, we can perform vectorized computations on them.

The `mpg$cty` column is in *miles/gallon*, let's convert to *km/l*.

```{r}
cty_km_l <- mpg$cty * 1.60934 / 3.78541 # miles/gallon to km/l
cty_km_l[1:5] # first five values
```

## Storing new columns

We can store the new column in the table by assigning it to a new column name.

```{r}
mpg$cty_km_l <- cty_km_l
mpg$cty_km_l[1:5]
```

We will see even better ways of doing this soon.

## Operations for Inference

R has many functions built in to perform **inference** (guessing about populations based on our sample).

* Hypothesis testing: state a hypothesis about a population parameter, decide if data are consistent with the hypothesis at given level of confidence (willingness to make a mistake)
* Estimation: guessing about population values, include uncertainty in the form of confidence intervals
* Prediction: guess about new data that has not yet been observed

## Wolf and Moose Populations on Isle Royale

The [Wolf-Moose Project] has been cataloging the size of wolf and moose populations on Isle Royale in Lake Superior for 60 years.

Here is a table of their data:

```{r}
wolf_moose <- read_csv("data/wolf_moose_yearly.csv")
dim(wolf_moose)
summary(wolf_moose)
```

We will treat these data as being a sample from a population of possible yearly observations that we could make.

## Names with invalid identifiers

Recall in R, our variables must start with a letter, `.`, or `_` and consistent of only those and numbers.

Notice that some of the colnames of the `wolf_moose` table do not adhere to that:

```{r}
wolf_moose |> colnames() |> tail(2)
```

We can access those using backticks around the names:

```{r}
wolf_moose$`ice bridges (0=none, 1 = present)` |> mean()
```

To make our life a little easier, let's change the last column name:

```{r}
colnames(wolf_moose)[15] <- "ice_bridge"
wolf_moose |> colnames() |> tail(2)
```

Notice the use **assigning to a function.** R has some special functions that let you do this. It is not automatic.

## Hypothesis Tests

Many of the built in test functions have `.test` suffixes on their name.

For example, if we wanted to test if the proportion of years (out of all possible years) that had an ice bridge between Isle Royale and the main land was 0.5 (against the alternative that is not 0.5), we can use `binom.test`:

```{r}
formalArgs(binom.test)
binom.test(sum(wolf_moose$ice_bridge), dim(wolf_moose)[1])
```

Many of the hypothesis tests also include confidence intervals for an estimate.

Student's t-test is `t.test`. Let's test the hypothesis that the moose population size is 1000 on average

```{r}
formalArgs(t.test) # need to get more info!
t.test(wolf_moose$moose, mu = 1000, conf.level = 0.99)
```

## Linear Regression

Recall that linear regression models the **conditional mean** of one variable ($Y$) as a linear function of other, conditioning variables ($x_1, x_2, \ldots$):

$$E(Y \mid X_1 = x_1, X_2 = x_2, \ldots) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \ldots$$

We can get estimates for the $\beta$ parameters using the `lm` function. The `lm` function uses R's **formula** interface:

```{r}
lm(wolves ~ moose + ice_bridge, data = wolf_moose)
```


To get the hypothesis tests for the model, we need to use the `summary` function:

```{r}
lm(wolves ~ moose + ice_bridge, data = wolf_moose) |> summary()
```

