---
title: "Stats 306: Lecture 0"
subtitle: "Introduction to Stats 306"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
```

## Stats 306: Introduction to Statistical Computing

This course introduces basic concepts in computer programming and statistical computing techniques as they are applied to data extraction and manipulation, statistical processing and visualization. It is aimed primarily at undergraduate majors and minors in Statistics.

*Enforced Pre-requisite: [DATASCI 101 or STATS 206 or STATS 250 OR STATS 280 OR STATS 412 or ECON 451 or IOE 265] AND prior or concurrent enrollment in [MATH 116 or MATH 121 or MATH 156 or MATH 176 or MATH 186 or any 200+ level math credit except for MATH 201].*

## Learning Objectives

By the end of this course, you will

* have a rigorous understanding of the structure of computer programs including
  * variables
  * functions
  * iteration
  * input/output
* have thorough familiarity with the R programming language and RStudio development environment
* be able to load, manipulate, and present data
* manage development projects using the Git version control package

## Overview of Course

* Lecture: introduce concepts and key R programming tools, frequent exercises
* Lab: Hands on programming experience with GSI assistance
* Homework: Independent learning in structured format
* Projects: Open ended opportunities to demonstrate abilities
* Quizzes: Quick checks of comprehension, maintain pace

## Example Exercise

Bring a laptop (or even a tablet!) to class:

We will frequently stop an engage in exercises. For example,

Add the values two and three:

```{r add23, exercise = TRUE}

```

```{r add23-hint}
2  + ...
```

```{r add23-solution}
2 + 3
```


## Grading

The course will be based on the following structure:

* Homework Assignments (40%) (9 assignments, no drops)
* Quizzes (20%) (13 quizzes, two lowest  dropped)
* Labs (5%) (two allowed absences)
* Project 1 (10%)
* Project 2 (25%) 

Late homework will not be accepted. Please email your GSI if you have circumstances requiring an extension.

## Working in Groups

You are encouraged to work together on **all class material** to discuss
solutions and debug. Each student must submit own work. Ask yourself, "Am I
avoiding learning something in my actions?" If yes, don't do it.

## Materials

All books are freely available through our library system or online:

* Our main text book is [R for Data Science (2ed)](https://r4ds.hadley.nz/) by Hadley Wickham, Mine Çetinkaya-Rundel, and Garett Grolemund
* Later in the semester, we will pull additional material from
    * [Advanced R](https://adv-r.hadley.nz/index.html)
    * [Mastering Shiny](https://mastering-shiny.org/)
* For learning git, we will use [Beginning Git and GitHub](https://search.lib.umich.edu/catalog/record/99187305231206381) by
Mariot Tsitoara.
* For statistical background, we suggest [Practical Statistics for Data Scientists](https://search.lib.umich.edu/catalog/record/99187344801606381)


## Student Hours and Help

* Instructor Zoom student hours TBA 
* GSI office hours: You may attend any GSI's hours
* Piazza on Canvas

## Other policies

Canvas includes more detailed policies on

>* Academic Misconduct
>* Classroom Culture of Care
>* Course Recordings
>* Disability Statement
>* Mental Health and Well-Being
>* Sexual Misconduct Policy

## Canvas

Visit Canvas....

## Software

* R: A statistical programming language
* RStudio (Cloud): A development environment for R
* Git: Version control and collaboration for software projects

  
## R, a statistical programming language

R is a *statistical programming language*:

* Statistical: designed for the manipulation and analysis of data
* Programming: perform analysis through collection of instructions
* Language: text based, as opposed to graphical, instructions

## Why use R?

* Purpose built for statistical analysis
* Large collection of add-on libraries containing cutting edge resources
* Being adopted by research, industry, government
* Language of choice for statistical teaching (courses, books)


## Using our Posit Cloud

* There is a main module: `stats306_winter_2024`. If you load this it will be a *temporary* copy. If you wish to retain any notes, you must **make it permanent**.
* If you make it permanent, you will need to update using `git` (more on this shortly)
* Labs and homeworks will be distributed as weekly modules. There is an example. This will automatically be made permanent.

Let's visit [Posit Cloud](https://posit.cloud/spaces/462616/content/all?sort=name_asc)

## What is Git?

* A **version control system** manages changes from multiple sources simultaneously.
* A reasonable analogy: Google Docs with track changes turned on
* Useful for groups working together to track changes to the software package and document why changes were made
* We will use it to disseminate homework assignments, labs, and lecture materials.

## How does git work?

* Creates a **repository** (a hidden directory/folder) to information about the history of files
* When you make changes, you may **add** them to a **staging area**
* When you like your staging area, you **commit** your changes.
* Git stores the **differences** ("diffs") from the previous version to your new version.
* You can **push** your changes to a shared repository for others to use
* You can **pull** from a shared repository and **merge** the changes of others.
* If there are **conflicts**, git will let you know and make you resolve them.
* If you ever need to see what changed, you have the entire **history** of all changes to look at.

## Command line interfaces

While git has graphical interfaces, we will teach using a **command line interface** (CLI) for git.

In an CLI you have an **interactive** prompt where you type a command. It might look something like this:

```
$ command -o --options argument1 argument2
```

* `$` is the prompt
* `command` is what we want to do (e.g., `git`)
* Optional flags have `-` or `--` in front and change behavior
* We can tell what do with `argument1` and `argument2`. Some commands have no arguments.

## Some basic CLI commands

* `cd`: Change directory. 
* `ls`: List files in current directory
* `pwd`: Print working directory
* `mkdir`: Make a new directory (folder)
* `rm -r`: Recursively delete directories and everything inside (use carefully!)

## Git and the CLI

When interacting with git, we will use commands like this:

```
$ git SUBCOMMAND -o --options ARG1 ARG2 ...
```

Some example subcommands will be `pull` to get updates, `log` to see the history, `add` to stage files, and `commit` to commit changes.

## Let git know who you are

```
$ git config --global user.name "YOUR NAME"
$ git config --global user.email "YOUR EMAIL"
```

## Getting a copy of the Stats 306 Repository

Materials for this course are hosted at **GitHub**, a popular website front end for git. Our materials are stored in repository on github.

If you want to get a copy (outside of Posit cloud) **clone** the repository:

```
$ cd LOCATION_YOU_WANT_TO_STORE_REPO
$ git clone https://github.com/markmfredrickson/stats306_spring_2024.git
```

## Getting future updates

When we make changes to the course materials (this will be frequent! We like to tinker!), you can get updates:

```
$ git pull
```

## Saving changes to our local repo

If you have made changes to your repository, you will be prompted to commit them before pullling:

```
$ git status
$ git add FILENAME
$ ... perhaps other files ...
$ git commit -m "a note explaining changes"
```

## Graphical interface

We teach the command line interface for git because

* It is convenient to show in slides
* It is transferable to any environment where you would use git
* It matches our recommended text book
* It exposes the full power of git

But, you can also use a graphical interface in Posit Cloud

## Posit Cloud and RStudio

You may encounter the name **RStudio**. 

This is the name a desktop version of the editor we are using (formerly it was called RStudio Cloud).

If you wish, you can [install R and RStudio](https://posit.co/download/rstudio-desktop/) on your own computer. 

You can get the materials using `git` as described above. You will need to download labs and HW from the assignments on Posit cloud.

## Final notes

**No lab this week**. Next activity is lecture on Tuesday.

Join the Stats 306 workspace on Posit (see Canvas). Try out the example module.

