---
suutitle: "Stats 306: Lecture 1"
title: "Starting R, RMarkdown, and Functions"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
```

## Opening up Stats 306 repository

Go to Posit Cloud and open up the `stats306_materials` repository.

## What's in our repository

>* Data sets will we use
>* Lab notebooks 
>* Lecture notes (like these)

All homework, labs, and lectures are **RMarkdown** documents.

## RMarkdown documents

RMarkdown documents blend **code** and **presentation**. 

General workflow:
* Open `.Rmd` file
* Run code in R chunks
* Edit text and code to answer prompts
* Knit to HTML file
* (Homework/projects) submit HTML files on Gradescope

## What is markdown?

* Markdown files are **plain text** files that have special interpretation.
* Came out of common conventions for adding expression to emails.
* They have a header (denoted by `---`) and a body. We can **render** them to HTML to view in any web browser.
* See the "Help -> Markdown Quick Ref" in RStudio for more details.

## RMarkdown

* RMarkdown extends basic markdown to allow embedding R code within the document.
* It addition to formatting text, you can include code **chunks**.
* When you **knit** the document, first the code is evaluated, then the results are added to the markdown document. Then that is turned into a web page.
* In addition to text and numeric output, plots and graphs are automatically inserted into the output.
* There are options to control how chunks are presented, cache computations to speed up knitting, etc.

## Chunks in Rmd files

All chunks have the following form:

<!-- The extra stuff is to get this print out as if
it were an R chunk, even though it is not really one. -->
<pre><code>```{r name, key = value, ...}
... R code goes here ...
<span>```</span></code></pre>

* The `name` is optional but helps when you get an error.
* The `key = value` controls the behavior of the chunk (optional). Example, `include = FALSE` to hide chunk.
* `CTRL-ALT-I` inserts a new chunk your RMarkdown document.

## Example 

Using an example homework file, we'll answer a few questions and prepare our HW for submission.

## R's Read-Evaluate-Print-Loop (REPL)

* The **Console** in R lets you interact with a running R process.
* When you type in in the console, R **reads** the command to make sure it is valid R code.
* R **evaluates** the valid code -- produce the result of the computation
* R **prints** the result back to the console.
* The process then repeats (**loop**)

## Trying out the REPL

We will use exercises in the slides like a REPL. Use the math operations `*`, `\`, `+`, and `-` to compute twice the difference of 100 and 18. You can use `()` to group calculations.

```{r repl, exercise = TRUE, exercise.lines = 1}

```

```{r repl-solution}
(100 - 18) * 2
```

The "Start Over" button will give you a fresh environment.

## Comments

We can include human readable text in R files and chunks using the `#` character.

This can make code easier to read:

```{r}
# The geometric mean of a set of numbers is defined as the exponentiation of the (arithmetic) mean of the log values
geo_mean <- function(x) {
  mean(log(x)) |> exp()
}

# Here is an example using the numbers to 1 to 10
geo_mean(1:10)
```

Comments can also be used for *documentation* (using a mini-language for documentation):
```{r}
#' Compute the geometric mean of a set of numbers.
#'
#' @param x A vector of values
#' @return The geometric mean of the input.
#' @example geom_mean(1:10)
geo_mean <- function(x) {
  mean(log(x)) |> exp()
}
```

We will also use comments for interacting with the Copilot AI assistant:

```{r}
# prompt: write a function that computes the geometric mean of a set of values
```


## Expressions and Statements

If you stopped a running R program at some point in time, everything about the program is called its **state**.

An **expression** is R code that, when run, provides a value (do not change state):

```{r eval = F}
3 + 4
```

Commands that change the **state** of the program are called **statements**, such as assignment:

```{r}
a <- 3 + 4
```

R runs statements one at time (each line is a statement) from top to bottom of the program or RMarkdown document.

(Technically `3 + 4` on a line is a statement because R will print out the value, but we will think of it as an expression.)

## Including Expresssions in Markdown

Suppose you need to calculate a value like $99 \times 87$. You could use R to help you calculate it

```{r}
99 * 87
```
Then copy it into a some text:

> The size of a rectangle with length 99 and width 87 is 8613.

Better way: Put the calculation right into your document!


> The size of a rectangle is &#96;r 99 * 87&#96;.


Results in:

> The size of a rectangle is `r 99 * 87`.

See the `lecture01.Rmd` file to see this in action.




## Variables and Assignment

A **variable** is a box to store data. We put the results of expressions into variables using **assignment**.
```{r}
a <- 7
b <- 12
c <- a + b
print(c)
```

Rules for variable names: 

* start with a letter,  `.`, or `_`. 
* followed by letters, numbers, `.`, `_`

You can include variables in your markdown to get their values.

> The value of &#96;c&#96; is &#96;r c&#96;.

> The value of `c` is `r c`.

## REPL and State

Suppose you entered the following at your R console:
```{r}
a <- 3 + 1
a <- 10 + 2
```
```{r repl-state-setup}
a <- 3 + 1
a <- 10 + 2
```
What would happen if you did the following?
```{r repl-state, exercise = TRUE}
print(a)
```

## The REPL and Chunks

When you **knit** an RMarkdown document, RStudio **starts a new process**.

Every time it encounters a **R chunk** it **evaluates** the R code inside and inserts the output into the document (default; can be changed). Like a behind the scenes REPL.

After running all code, it turns the markdown into a web page (`example-homework.Rmd` => `example-homework.html`)

## Example of chunks and state

(Feel free to look at `lecture01.Rmd` right now).

Here is a code chunk:
```{r}
classname <- "Stats 306"
```
Here is some intervening text and a second chunk:
```{r}
print(classname)
```

## Order is important

Because **statements** change the context, order of statements is very important. Compare
```{r}
a <- 7
print(a)
a <- 2
```

```{r}
a <- 7
a <- 2
print(a)
```

## Working with code chunks, debugging hint

You can send a line of code to the console using `CTRL-Enter`.

You can run all of the code in a chunk at once using the green arrow or `CTRL-SHIFT-Enter`.

If you run into a problem, use the `Session -> Restart R` to clean up, and the use "Run all chunks above" (`CTRL-ALT-SHIFT-p`). Then try the chunk again.

## What can go in variables?

R has several different **data types**. Some common types:

* **Numeric**: `7`, `3.22`, `9.1e2`
* **Strings**: `"Hello"` or `'Hello'` (also called "character vectors")
* **Logicals**: `TRUE` and `FALSE`
* We'll see more data types going forward
* These are also referred to as the **class** of the data.

## Why are variables useful?

* Avoid repeatedly typing in the same thing
* Save the results of computations
* Avoid recomputing same values
* Separate what we are computing from any particular input (**abstraction**)

## Exercise: Using variables

Find the difference between 193 and 7. What is this value divided by two? What is this value divided by three? (Only compute the difference once.)

```{r using-variables, exercise = TRUE, exercise.lines = 5}

```

```{r using-variables-solution}
d <- 193 - 7
d / 2 # will automatically print out
d / 3
```


## Style point: descriptive variable names

* So far we have used names like `a` and `b`.
* For large programs, having variable names that describe what they hold is much better
* Compare `a` to `number_of_seconds_in_a_day`
* In RStudio/Posit, it's just as easy to type with `TAB` completion

## Functions 

A key tool in any programming language is a **function**.

* R functions behave (mostly) like mathematical functions, e.g. $f(x) = x^2$.
* They take **arguments** and return **values**
* Encapsulate code (mini-programs)
* Make it easy to repeat operations

## Functions in R

Once we have data stored in a variable, one of the main ways to use it is to **call a function.**

```{r, eval = FALSE}
return_value <- function_name(argument1, argument2, named_argument = additional_data, ...)
```

* Like math functions, we use the syntax `f()` to run the function
* All **arguments** are **passed** to the function inside of the `()` separated by commas
* Arguments can be passed in **positional** order or
* Passed by **name**
* Some arguments are **optional** when they have a default value
* `...` means "any number of additional arguments"
* `return_value` is the output of the function


## Converting from degrees to radians

Recall two ways to measure an angle:

>* Degrees: 0 to 360
>* Radians: 0 to 2$\pi$

Exercise: compute the radian equivalent of 45 degrees (use a good variable name!). Then use the `sin(your_variable_name)` function to compute the sine of 45 degrees. (Hint: there is a built in variable named `pi`.)

```{r sin45, exercise = TRUE}

```

```{r sin45-solution}
rad45 <- (45 / 360) * 2 * pi
sin(rad45)
```


## Function evaluation

In R, when we call a function, first the **arguments** are evaluated, then the function is called.

```{r}
sin(360/45 * 2 * pi) 
```
Equivalent to:
```{r}
value1 <- 360/45
value2 <- value1 * 2
value3 <- value2 * pi
sin(value3)
```

## Avoiding single use variables

Sometimes we only need to use something once, so taking the time to come up with a variable name is unnecessary. 

Instead we can use **pipes** to pass the output of one function to the **first argument** of a second function.

```{r, eval = FALSE}
g(f()) # f() evaluates first, then the result goes to g.

# equivalently
value <- f()
g(value)

# pipes
f() |> g()
```

## Average BMI for women

The `women` dataset is built into R as an example we can play with. It contains the average weight given height for a sample of women with heights between 58 and 72 inches.

```{r}
head(women) # first 6 rows of the table
```

Suppose we want to change the units into metric:

```{r}
women2 <- transmute(women, height_m = height * 0.0254, weight_kg = weight * 2.2)
head(women2)
```

Then we want to compute the average BMI:

$$\text{BMI} = \frac{\text{Weight (kg)}}{(\text{Height (m)})^2}$$

```{r}
women3 <- mutate(women2, bmi = weight_kg / (height_m)^2)
head(women3)
```

Finally, let's plot BMI as a function of height:

```{r}
ggplot(women3, aes(x = height_m, y = bmi)) + geom_line()
```


## Using pipes

In the previous example, we didn't use the `women2` variable except to create `women3`. And we only graphed `women3`. Let's use pipes to eliminate those variables.

```{r}
women |> # style point: create a new line after the pipe character
  transmute(height_m = height * 0.0254, weight_kg = weight * 2.2) |>
  mutate(bmi = weight_kg / (height_m)^2) |>
  ggplot(aes(x = height_m, y = bmi)) + geom_line()
```

## Pipe exercise

Recall that logarithms answer the question:

$$\log_b(a) = c \text{ such that } b^c = a$$
As a consequence, 

$$log_b(b^a) = a$$
Verify that the `log` and `exp` functions are inverses of each other (using base $e \approx 2.71$) using pipes. Rewrite the following to use pipes.

```{r log-exp-ex, exercise = TRUE}
a <- log(7)
b <- exp(a)
print(b)
```


## R's assignment operators

R has two ways to store data in variables:

```{r}
variable_arrow <- 7
variable_equal = 10
variable_arrow + variable_equal
```

I like the arrow because it can be reversed
```{r}
7 -> variable_reversed
pi * variable_reversed
pi |> sqrt() -> sqrt_pi
sqrt_pi
```

## Alternative pipe operator

* There is also an alternative pipe operator `%>%`.
* It is (almost) equivalent to `|>` with `|>` being a little more strict about using parentheses.
* The choice is mostly a matter of muscle memory.

## Some special values

In addition to `pi`, R has some other built in variables (see help page for `Constants`).

Also several special data values

* `NA` means "missing". Including a `NA` value in a calculation will usually make the result also `NA`.
* `NULL` means "not set yet" and is often used for optional arguments to functions
* `NaN` means something was divided by zero or other mathematical impossibility.

What will happen with this code and why?

```{r na-prop, exercise = TRUE}
angle_degrees <- NA
angle_radians <- 2 * pi * angle_degrees / 360
sin(angle_radians)
```

## Summary

* RMarkdown documents: text and code chunks
* R and the REPL
* Variables: boxes to hold data
* Functions: little programs that operate on data
* Assignment operators
* Pipes


