---
title: "Data Science 306: Lecture 8"
subtitle: "Wrapping up functions"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(lubridate) # install.packages("lubridate") if you don't have this
library(Lahman) # install.packages("Lahman") if you don't have this
```

## Review

* From duplicated to clean code:
  * Common patterns in function
  * Use of `across`
  * Lists of functions as arguments
* Return values: only one but can package up in vectors, lists, and tables
* Predicates: functions that return `TRUE` or `FALSE`; use in `select`
* Anonymous functions, tidyverse short hand notation `~ (.x - mean(.x))^2 / (length(.x) - 1)`

## The `Lahman` package

Sean Lahman maintains an extremely [detailed database of American baseball
statistics](https://www.seanlahman.com/baseball-archive/statistics). The
`Lahman` package conveniently imports this data for us.

```{r}
library(Lahman)
data(package = "Lahman")
dim(Batting)
colnames(Batting)
```

We can look up the description of the columns using `?Batting` or the help menu.

## Getting in the Lahman data

Let's investigate some successful players over the history of baseball:

```{r}
range(Batting$yearID)
group_by(Batting, playerID) |> 
  summarize(across(c(G, HR, RBI), sum)) |>
  arrange(desc(HR)) |> head(5)
```

## From IDs to full names

The `People` table has a `playerID` column. Does this uniquely identify players?

```{r}
dim(People)
summarize(People, n_distinct(playerID))
```

Looking up players: 
```{r}
lookupPlayer <- function(pid) {
  filter(People, playerID %in% pid) # result will be a table
}

lookupPlayer(c("aaronha01", "ruthba01")) |> select(nameFirst, nameLast)
```

## Writing functions that take columns as arguments

Many of the tidyverse functions take columns names as arguments. 

This functionality is non-standard for R. Usually arguments are evaluated *before* calling a function.

```{r eval = FALSE}
# Run this in your console to see the error
f <- function(tbl, column) {
  mutate(tbl, column * 2)
}

f(starwars, mass)
```

This will cause an error:

```
Error in `mutate()`:
ℹ In argument: `column * 2`.
Caused by error:
! object 'mass' not found
```

Such **non-standard evaluation** is quite helpful when doing data analysis directly, but makes writing functions a little more difficult.

## Curly brace syntax

The error is caused by the fact that "mass" was not a variable. To indicate that we should not interpret it as a variable, and instead think of it as a column name we use `{{column}}` syntax:


```{r}
# Run this in your console to see the error
f <- function(tbl, column) {
  mutate(tbl, {{column}} * 2)
}

f(starwars, mass) |> select(mass, `mass * 2`)
```



## Successful players (batting)

Which players had the most home runs? Games played? Stolen bases?

We want to compute the same thing several times, we'll use curly brace synatax to help:

```{r}
top_players <- function(stat) {
  group_by(Batting, playerID) |>
    summarize(stat_sum = sum({{stat}})) |>
    arrange(desc(stat_sum)) |>
    head(1) |>
    mutate(lookupPlayer(playerID)) |>
    select(playerID, stat_sum, nameFirst, nameLast)
} 

# stack tables
bind_rows(
  top_players(HR),
  top_players(G),
  top_players(SB))
```

## Variable number of arguments

Recall our template for creating a function in R:

```{r, eval = FALSE}
fname <- function(required_arg, optiona_arg = 123, ...) {
  # function body
}
```

The `...` means "any number of additional arguments"

## Example of variable arguments

```{r}
sum_summary <- function(...) {
  select(...) |> summarize(across(everything(), sum))
}

sum_summary(Batting, HR, G, SB)
```

Notice how all three arguments get passed to `select` automatically.


## Exercise: Creating URLS

The `paste` function will take more than one item to turn into a string using its `...` argument:

```{r}
paste("Welcome", "to", "DataSci306", "Introduction", "to", "Statistical", "Programming")
```

Write a function that takes a `...` argument and passes it along to `paste`. Set the `sep = "+"` when calling `paste`.

```{r exdots, exercise = TRUE}

##plus_paste("Welcome", "to", "DataSci306", "Introduction", "to", "Statistical", "Programming")

```

## Using anonymous functions to pass arguments to called functions

We have seen `na.rm = TRUE` pop up from time to time.

```{r}
d <- tibble(x = c(1, 10, NA, 7), y = c(1000, 2000, 10000, NA))
d
```

What happens if we try to summarize using `mean`?

```{r}
summarize(d, across(everything(), mean))
```

We can make a version of the `mean` function with `na.rm = TRUE` using an *anonymous* function:

```{r}
summarize(d, across(everything(), ~ mean(.x,  na.rm = TRUE)))
```

## Fixing a previous function

```{r}
sum_summary <- function(...) {
  select(...) |> summarize(across(everything(), ~ sum(.x, na.rm = TRUE)))
}

sum_summary(Batting, HR, G, SB)
```

## Conditional evaluation, short circuiting

We have used `if_else` to make choices for an entire vector. We often need **conditional program flow** in functions.

```{r}
my_median <- function(x) {
  x_sorted <- sort(x)
  n <- length(x)
  if (n %% 2 == 1) {
    ## odd number of observations
    return(x_sorted[(n + 1)/2])
  } else {
    return(mean(x_sorted[c(n/2, n/2 + 1)]))
  }
  
  return(Inf) # this line will never run
}

values_even <- c(19, 2, 4, 11)
values_odd  <- c(19, 2, 4, 11, 9)

my_median(values_even)
my_median(values_odd)
```

## Careful about what you put in your `if` statements:

```{r eval = FALSE}

if (values_even > 3) {
  ... code might assume all values are larger than 3 ...
}
```
Would give the error message:
```
Error in if (values > 3) { : the condition has length > 1
```

Solution: use the `any` and `all` functions

## Exercise: Safe variance

Recall the usual (unbiased) estimator for variance:

$$\frac{\sum_{i = 1}^n (X_i - \bar X)^2}{n - 1}$$

If there is only one unit in the sample, this estimator is undefined. Instead, let's estimate the variance as 0 if we only have one observation.

Write a function to estimate variance in this way. Demonstrate on a vector of size 1 and a longer vector. Use `var` in the case where $n > 1$.

```{r safevar, exercise = TRUE}

```

## Combining more than one condition

We can use `&` for "and" and `|` for "or". We can use `!` to negate.

Here's an example where we can only set one option for a function:

```{r}
food_order <- function(entree, side = NULL, dessert = NULL) {
  if (!is.null(side) & !is.null(dessert)) {
    return("You may only order one side or dessert")
  } 
  
  extra <- 
    if (!is.null(side)) {
      paste("with a side of", side)
    } else if (!is.null(dessert)) {
      paste("with a dessert of", dessert)
    } else {
      # both NULL <=> nothing extra ordered
      "with no other items"
    }

  paste("You ordered a main dish of", entree, extra)
 
}

food_order("soup")
food_order("soup", side = "salad")
food_order("soup", dessert = "cake")
food_order("soup", "salad", "cake")
```

## Environments and functions

An **environment** in R is all defined variables and their values (we can think of it like a list). 

You have a **global** environment for your R session, and R packages and functions have their own environments.

```{r}
x <- 3
f <- function(y) {
  x <- 2
  y + x
}
f(2)

```

## Lexical scoping

Environments are **nested**. If we request a variable not in the current environment, we go up the chain. We call this lexical scoping.

```{r}

x <- 3
f <- function(y) {
  y + x
}
f(2)
```
## Small aside: `pull` function

Sometimes we get tables, but we actually just want a column.

```{r}
## we want to get the memory for the laptop:
tbl <- tibble(computer = c("Desktop", "Laptop"), RAM = c(16, 8))
(lap_mem <- filter(tbl, computer == "Laptop") |> select(RAM))
```

We can get this value by pulling out the RAM column:


```{r}
pull(lap_mem, RAM)
```

## Using lexical scope in a function

This can be convenient if we want to calculate a quantity and then re-use it when summarizing or mutating.

Barry Bonds has hit the most homeruns. 
```{r}
filter(Batting, playerID == "bondsba01") |>
  summarize(s = sum(HR)) |> 
  pull(s) -> bb_hr

bb_hr
```

```{r}
at_least_bb <- function(x) { 
  x > bb_hr # x is argument, bb_hr is from environment
}
```


How many players hit at least as many hits, doubles, or triples as Bonds hit HRs?
```{r}
group_by(Batting, playerID) |>
  summarize(across(c(H, X2B, X3B), sum)) |>
  transmute(across(c(H, X2B, X3B), at_least_bb)) |>
  summarize(across(everything(), sum))
```

## Plotting hits

```{r}
group_by(Batting, playerID) |>
  summarize(H = sum(H)) |>
  ggplot(aes(x = H)) + geom_histogram() + geom_vline(xintercept = bb_hr)
```

## Exercise

Which columns contain a value greater than 100?

To find out, first, select only non-factor columns (`is.factor` will be helpful), then select the columns that have a value of at least 100 somewhere in column. Use `colnames()` to list the column names.

Then find the opposite. What columns do not have values greater than 100? Has anyone ever been hit by pitches more than 100 times in a season (column HBP)

```{r selectexercise, exercise = TRUE}

```

```{r selectexercise-solution}
select(Batting, where(~ !is.factor(.x))) |> select(where(~ max(.x, na.rm = TRUE) > 100)) |> colnames()

select(Batting, where(~ !is.factor(.x))) |> select(where(~ max(.x, na.rm = TRUE) <= 100)) |> colnames()

```
