---
title: "Stats 306: Lecture 2"
subtitle: "Vectors, Lists, and Tables"
output: 
  learnr::tutorial:
    progressive: true
    css: css/lecture.css
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
```

## Statistical programming

Recall that R is a **statistical** programming language. In statistics we often work with:

>* **Samples**, composed of individual units
>* **Variables**, something measured on each unit in the sample

Tabular data: columns are variables, rows are observations

## Lists and Vectors

* A **vector** is an ordered collection of the same type of data. Created using the `c` function (concatenate).
* A **list** is an ordered collection of (potentially) different types of data. Created using the `list` function.
* Both can have **names** to indicate individual items.
* Lists can serve a similar role to dictionaries and hash tables in other languages.

## Vectors, basic

```{r}
phone_digits <- c(8, 6, 7, 4, 3, 0, 9)
phone_plus_area_code <- c(7, 3, 4, phone_digits)
length(phone_plus_area_code)
sum(phone_plus_area_code)
```

## Vectors, indexing

The `runif` function will create randomly generated values in the range (0, 1).

```{r}
random_u01 <- runif(10) # create 10 values
print(random_u01)
random_u01[1] # first item
random_u01[10] # last item
random_u01[-1] # all but first
random_u01[1:3] # first three
random_u01[3:6] # a "slice" in the middle
random_u01[c(1, 3, 7, 1)] # arbitrary choices
```


## Exercise

Pull out the last three items in the vectors `letters`. A little help to get you started:

```{r lettersxyz, exercise = TRUE}
length(letters)
```


```{r lettersxyz-solution}
letters[c(24, 25, 26)]
letters[-(1:23)]
```

## Relational Operators

```{r}
x <- 1
y <- 2

x < y
x > y
x <= y
x >= y
x == y
x != y

!(x == y) # negation
```

## Vectors, subsetting based on qualities

```{r}
less_than_0.5 <- random_u01 < 0.5
print(less_than_0.5)
random_u01[less_than_0.5]
keep(random_u01, random_u01 < 0.5)
```

## Exercise

Use the following hint to keep only the even numbers.

```{r}
(7 %% 2) == 0 
(8 %% 2) == 0
```


```{r onlyodd, exercise = TRUE}
x <- 1:10
```

```{r onlyodd-solution}
x <- 1:10
keep(x, (x %% 2) == 0)
```


## Lists, heterogenous collections

Vectors must be the same type, lists can be **heterogeneous**:
```{r}
list("string", 7, 1:4)
```

## Named collections

Vectors can have names (`c(first = 1, second = 22, third = 333)`), but names for lists are more common.
```{r}
named_list <- list(a_string = "A string", a_vector = 1:4)
named_list[["a_string"]] # note double brackets
named_list$a_vector
```

## Tables: Lists of Vectors

* In a **sample**, we suppose we have $k$ variables measured for each of $n$ units.
* For each variable, place the $n$ values in a vector.
* Collect all such vectors into a list of length $k$.
* Result: $n \times k$ table

## Tables in R

There are a few different types of tables in R (`data.frame`, `tibble`). We will treat them interchangeably for the most part.

```{r}
class(mpg)
dim(mpg)
colnames(mpg)
head(mpg)
```

## Accessing individual columns

Recall, the `$` operator lets us get items from lists:

```{r}
mpg$hwy |> mean() # equiv to mean(mpg$hwy)
```

## Functions for basic descriptive statistics

```{r}
sum(mpg$hwy)
mean(mpg$hwy)
var(mpg$hwy)
sd(mpg$hwy) 
median(mpg$hwy)
IQR(mpg$hwy)
quantile(mpg$hwy, c(0.25, 0.75))
```

## Exercise

What is the average `year` of the cars in `mpg`? What is the standard deviation of the city mileage (`cty`)?

```{r mpgsummary, exercise = TRUE}

```

```{r}
mean(mpg$year)
mpg$cty |> sd()
```



## Dealing with missingness

Here's a vector with some missing values
```{r}
v <- c(1, NA, 2, 10, 3, NA)
mean(v)
```

To delete the missing values we have some options:
```{r}
mean(v, na.rm = TRUE)
na.omit(v) |> mean()
v[!is.na(v)] |> mean()
```

## TODO

* Logical data types and comparisons?
* More on tables, really basic summaries
* Other gotchas
* **Vectorized computations**